Parsing Postfix log files

### 2008-03-03 doc/logparser.tex
Changes after proof-reading.

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm, explaining the difficulties
    encountered and the solutions developed, with reference to an
    implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\url{http://www.postfix.org/}} Further complication arises because the set
of rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every \RBL{} returns a different
explanatory message; policy servers may log different messages depending on
the characteristics of the connection; there are many ways in which the log
lines may differ between servers, even within the same organisation:
servers may be configured differently, or running different version of
Postfix.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database for later use.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying regular high volume uses (e.g.\ customer newsletters) and
restricting those uses to off-peak times; detecting virus outbreaks which
propagate via email; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in \sectionref{database schema}, explaining in detail the tables
used for storing the data gleaned from the logs and the table that stores
the rules.

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are discussed, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, with reference to the
graphs in appendix~\refwithpage{graphs}.  The final section is a
description of using the tools provided with the parser to generate a
\regex{} from related log lines for new rules.

Section~\ref{parsing-algorithm} contains the core of the paper, describing
a naive parsing algorithm and the complications encountered which shaped
the full algorithm.  A flow chart and a discussion of the emergent
behaviour exhibited by the algorithm accompanies a comprehensive
explanation of the different stages of the algorithm.  The actions taken
during execution of the algorithm are described, followed by the process of
adding a new action.  The section concludes with an in-depth description of
the further complications and their solutions which complete the algorithm.

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 1000 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines are parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

Section~\ref{conclusion} contains the paper's conclusion.

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency}, rule efficiency, and explains
the irregularities observed in these graphs.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP{}\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix anti-spam restrictions and policy servers.  The assumptions made in
designing and implementing the parser are explained, as are the conventions
used in this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

A short example of the optimisation possible using data from the database
is which Postfix restrictions reject the highest number of mails:

@@

If the database supports sub-selects percentages can be
obtained:

@@

\SQL{} note: $||$ is the concatenation operator in SQLite3; if the database
containing the extracted data does not support this syntax, then simply
remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
the same, just slightly less visually pleasing.

Another example is determining which restrictions are not effective: this
example shows which restrictions had fewer than 100 hits on the last log
file parsed.

@@

    \item The majority of programming languages have existing code
        available allowing access to databases, so the majority of
        languages will be able to access the results obtained by running
        the parser.  If this parser was written to be utilised as shared
        code there would be little or no difficulty in using it when
        writing in the programing language it is implemented in, but every
        other programing language wishing to utilise it would require an
        interface layer to be written.  This requirement would drastically
        reduce the viability of any project wishing to build upon the work
        of this project.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item Other tables can be added to the database, e.g.\ to cache
                historical data.

@@

            \item If the database supports it, access can be granted on a
                fine-grained basis so that the finance department can
                produce invoices, the helpdesk can run limited queries as
                part of dealing with support calls, and the administrators
                have full access to the data.

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of \SQL{}, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  A good introduction to \SQL{} can be found
        at~\cite{sql-for-web-nerds}, others
        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.  Depending on the
        database in use there may be tools available which reduce or remove
        the requirement to know \SQL{}; for SQLite (the default database
        used by the implementation) there are several
        available~\cite{sqlite-guis}.

@@

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see \sectionref{rule characteristics} for details.)

@@

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}, the following is a brief overview.

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

check\_client\_access checks whether the \IP{} address of the connected
client matches the left hand side of each line in the file and returns the
right hand side of the first matching line.  Other restrictions determine
their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
address of the client is present in the listed \RBL{}, rejecting the
command if the client is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be used as exceptions to more general cases.  When the result is
the name of another restriction Postfix will evaluate the new restriction,
allowing restrictions to be chosen based on the client \IP{} address, HELO
hostname, sender address, recipient address, etc.  The administrator can
define new restrictions as a list of existing restrictions, allowing
arbitrarily long and complex sequences of lookups and restrictions.

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each \SMTP{} command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction}
        records specify which mail servers are allowed to send mail
        claiming to be from a particular domain.  The intention is to
        reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and joe
        jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote \IP{}
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

@@

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is too high
        the mail is rejected.

@@

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).  On a well run system it is extremely unlikely that any of
        these problems will arise, though it is of course possible,
        particularly when undergoing a deluge of spam.

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time; the source code for
Postfix 2.4.6 is 16MB.} there are advantages to treating Postfix as a black
box while developing the parser:

@@

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling interferes with it.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that logs produced by Postfix are not fixed; they vary widely
from host to host, depending on the set of restrictions chosen by the
administrator.  With this in mind, one of the parser's design aims was to
make adding new rules as easy as possible, to enable administrators to
properly parse their log lines.  The main design decision which enables
this is the separation of rules from the parsing algorithm, and secondly
the division of the algorithm into discrete actions.

\label{why separate rules and algorithm}

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the algorithm
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both
algorithm and rules.  Decoupling also creates a clear separation of
functionality: rules handle low level details of identifying log lines and
extracting data from a line, whereas the algorithm handles the higher level
details of following the path a mail takes through Postfix, assembling the
required data before storing it, dealing with complications arising, etc.

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
complication which occurs only four times in \numberOFlogFILES{} log files
tested.} Requiring changes to the parsing algorithm also complicates
upgrades, as the changes must be preserved during the upgrade, and may
clash with changes made by the developer.  \parsername{} allows the user to
add new rules to the database without changing the parsing algorithm,
unless the new log lines to be parsed require functionality not already
provided by the algorithm.  If the new log lines do require new
functionality, new actions can be added to the parser without modifying
existing actions or other parts of the algorithm; only in the rare case
that the new actions require support from other sections of the code will
more extensive changes be required.

@@

@@

The assumptions which must be satisfied for the parser to work correctly
were discussed; the conventions used in the paper were listed; a brief
comparison of this project against other Postfix parsers was provided,
followed by a review of the previously published literature.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in \sectionref{rules} but the fields are covered here.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
\sectionref{connections table}.  The results table has one or more rows per
row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
\sectionref{results table}.

@@

Rules are discussed in detain in \sectionref{rules}, but the rules table is
covered here.  Rules are created by the user, not the parser, and will not
be modified by the parser (except for the hits and hits\_total fields).
Rules parse the individual lines, telling the parser which fields to
extract and what action to take for that line.

@@

@@

    \item [restriction\_name] The restriction which caused the mail to be
        rejected.  Only applicable to rules which have a result of
        \texttt{rejected}, other rules will have an empty string.

    \item [postfix\_action] This is the action Postfix must have taken to
        generate this line, with two exceptions:

@@

@@

@@

        The following keywords are expanded (full explanations can be found
        in the source code):

>>>>

    Parsing Postfix logs is much more difficult than it first appears, but
    it \textit{is\/} possible to achieve a high degree of accuracy in
    understanding the logs, and thus accuracy in reconstructing the actions
    taken by Postfix to generate the logs.  This paper describes the
    creation of a parser, documenting the parsing algorithm, explaining the
    difficulties encountered and the solutions developed, with reference to
    an implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each log line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.\footnote{A
history of all changes made to Postfix is distributed with the source code,
available from \url{http://www.postfix.org/}} Further complication arises
because the set of rejection messages is not fixed: new messages can be
added by the administrator with custom checks; every \RBL{} returns a
different explanatory message; policy servers may log different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ between servers, even within the same
organisation --- servers may be configured differently, or running
different version of Postfix.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  There are numerous other uses for such data:
improving server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in section~\refwithpage{database schema}, explaining in detail the tables
used for storing the data gleaned from the logs and the table that stores
the rules.

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are discussed, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, with reference to the
graphs in appendix~\refwithpage{graphs}.  The final section is a
description of using the tools provided with the parser to generate new
rules (specifically the \regex{} in each rule) from unparsed log lines.

Section~\ref{parsing-algorithm} contains the core of the paper, describing
a naive parsing algorithm and the complications initially encountered which
shaped the full algorithm.  A flow chart and a discussion of the emergent
behaviour exhibited by the algorithm accompanies a comprehensive
explanation of the different stages of the algorithm.  The actions taken
during execution of the algorithm are described, followed by the process of
adding a new action.  The section concludes with an in-depth description
of the further complications discovered and their solutions which complete
the algorithm.

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 1000 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.  A random sampling of log lines was parsed, and the correctness
of the results manually verified, as part of determining the coverage of
the parser.

@@

Section~\ref{conclusion} contains the paper's conclusion, describing the
results of the research, design and implementation of the parser.

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency} (rule efficiency), and explains
the irregularities observed in these graphs.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
paper, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

A short example of the optimisation possible using data from the database
is determining which Postfix restrictions reject the highest number of
mails:

@@

If the database supports sub-selects percentages can be obtained:
\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
the database containing the extracted data does not support this syntax,
then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
results will be the same, just slightly less visually pleasing.}

@@

Another example is determining which restrictions are not effective: this
shows which restrictions had fewer than 100 hits on the last log file
parsed.

@@

    \item The majority of programming languages have existing code
        available providing access to databases, so the majority of
        languages will be able to access the results obtained by the
        parser.  If this parser was written to be utilised as shared code
        there would be little or no difficulty in using it when writing in
        the programing language it is implemented in, but every other
        programing language wishing to utilise it would require an
        interface layer to be written.  This requirement would drastically
        reduce the viability of any project wishing to build upon the work
        of this project.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item Other tables can be added to the database, e.g.\ to cache
                historical, summary or computed data.

@@

            \item If the database supports it, access can be granted on a
                fine-grained basis so that the finance department can
                produce invoices, the helpdesk can run limited queries as
                part of dealing with support calls, and the administrators
                can have full access to the data.

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
        every vendor implements a different dialect of \SQL{}, the basics
        are the same everywhere (analogous to the overall similarities and
        minor differences between Irish English, British English, American
        English and Australian English).  Depending on the database in use
        there may be tools available which reduce or remove the requirement
        to know \SQL{}; for SQLite (the default database used by the
        implementation) there are several available~\cite{sqlite-guis}.

@@

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.  There are many spam and virus
sending programs which are this deficient --- unfortunately others
(particularly newer programs) were written by competent programmers, or
utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see \sectionref{rule characteristics} for details.)  This design has
positive security implications also: those components which interact with
other hosts are not privileged, and those components which are privileged
do not interact with other hosts.

@@

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}, the following is a brief overview
only.

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a benefit of this delay is that Postfix has
more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

The restriction \texttt{check\_client\_access} checks whether the \IP{}
address of the connected client is found in the file and returned the
associated action if found; the method of searching the file is dependant
on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
address of the client is present in the \RBL{} \texttt{rbl.example.com},
rejecting the command if the client is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address, etc.  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each \SMTP{} command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  For example some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender address, recipient address,
        remote \IP{} address) is unknown; on second and subsequent delivery
        attempts from that triple the mail will be accepted.  The
        assumption is that maintaining a list of failed addresses and
        retrying after a temporary failure is uneconomical for a spam
        sender, but that a legitimate mail server must retry.  Sadly spam
        senders are using increasingly complex and well written programs to
        distribute spam, frequently using an \ISP{} provided \SMTP{} server
        from a compromised machine on the \ISP{}'s network.  Greylisting
        will slowly become less useful, but it does block a large
        percentage of spam mail at the moment; the most effective
        restrictions over the \numberOFlogFILES{} log files used in testing
        the parser are shown in table~\refwithpage{Summary of rejections}.
        Greylisting is obviously worth using, at least at the moment,
        particularly when you factor in Greylisting's position as the final
        restriction which a mail must overcome:\footnote{Greylisting is the
        final restriction a mail must overcome in the configuration used on
        the mail server the logs were obtained from; an administrator is
        free to use Greylisting at whichever position in the restriction
        list they feel is most appropriate for their mail system.}
        Greylisting only takes effect for mails which have passed every
        other restriction.

@@

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is too high
        the mail is rejected.  Postfix's restrictions are binary, and the
        administrator must manually whitelist clients if they are to bypass
        a restriction; using a threshold which requires hitting several
        restrictions frees the administrator from whitelisting clients
        which fall foul of one restriction only.

@@

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).  On a well run system it is extremely unlikely that any of
        these problems will arise, though it is of course possible,
        particularly when suffering a deluge of spam or a mail loop.

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.  There are a
        number of heuristics used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time; the source code for
Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
box while developing the parser:

@@

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling interferes with logging and other processing.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that logs produced by Postfix are not fixed; they vary widely
from host to host, depending on the set of restrictions chosen by the
administrator.  With this in mind, one of the parser's design aims was to
make adding new rules as easy as possible, to enable administrators to
properly parse their log lines.  To enable this the parser is divided into
three parts:

    \item [rules] These match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines and actions.  Rules are described
        in detail in \sectionref{rules}.

    \item [actions] Are invoked to deal with a log line once the line has
        been identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] Is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting.

It may help to think of the rules and actions as components which plug into
the framework.

\label{why separate rules, actions and framework?}

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both actions
and rules.  Decoupling also creates a clear separation of functionality:
rules handle low level details of identifying log lines and extracting data
from a line, whereas the actions handle the higher level details of
following the path a mail takes through Postfix, assembling the required
data before storing it, dealing with complications arising, etc.

Decoupling the actions from the framework simplifies both framework and
actions: the framework provides services to the actions, and doesn't need
to deal with the complications which arise, or the task of reconstructing a
mail's journey through Postfix; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately reconstructing each mail's journey through Postfix and dealing
with the complications described in \sectionref{additional complications}.

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}.  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

@@

    \item Assumptions made when designing and developing the parser.

    \item A description of the parser's novel three section design.

    \item Conventions used in this document.

    \item A brief comparison of this parser versus other Postfix parsers.

    \item A review of the literature previously published in this area.

@@

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in \sectionref{rules} but the fields are covered here
in \sectionref{rule attributes}.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in
\sectionref{connections table}.  The results table has one or more rows per
row in the connections table; the fields will be covered in detail in
\sectionref{results table}.

@@

Rules are discussed in detail in \sectionref{rules}, but the structure of
the rules table is covered here.  Rules are created by the user, not the
parser, and will not be modified by the parser (except for the hits and
hits\_total fields).  Rules parse the individual lines, extracting data to
be saved in the connections and results tables, and specifying action to
take for that line.

@@

    \item [id] A unique identifier for each rule which other tables can use
        to refer to a specific rule.

@@

    \item [restriction\_name] The restriction which caused the mail to be
        rejected.  Only applicable to rules which have an action of
        \texttt{REJECTION}, other rules should have an empty string.

    \item [postfix\_action] This is the action Postfix must have taken to
        generate this line.  This field is mostly ignored, but two values
        (IGNORED and INFO) have special meaning, as described below in the
        list of typical values.\label{postfix_action}

@@

            \item [ACCEPTED] Postfix has accepted a mail, and will
                subsequently attempt to deliver it.

            \item [BOUNCED] The mail has bounced, due to a mail loop,
                delivery failure, or five day timeout.

            \item [DELETED] The mail was deleted from the queue by an
                administrator.

            \item [DISCARDED] Postfix discarded the mail it was in the
                process of accepting, because it was either larger than the
                limit set by the administrator, or the client timed out or
                disconnected.

            \item [EXPIRED] The mail has been in the queue for five
                days\footnote{As with much of Postfix's behaviour, this is
                the default value but can be changed by the administrator
                if they choose.} without successful delivery.  A bounce
                mail will be generated and sent to the sender address.

@@

            \item [POSTFIX\_RELOAD] The administrator has caused Postfix to
                start or stop, and all existing \daemon{smtpd} processes
                will be terminated.

            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
                see~\cite{postfix-cleanup} for details of the processing
                performed by \daemon{cleanup}.

            \item [REJECTED] Postfix rejected a command from the remote
                client, causing at least one recipient to be rejected.

            \item [SENT] Postfix has successfully sent a mail.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in \sectionref{rule efficiency}, with the impact of
        compiling and caching \regexes{} covered in \sectionref{Caching
        each regex}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted from the log line.  The format is:
        \newline \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
        \newline i.e.\ semi-colon or comma separated assignment statements,
        with the variable name on the left and the matching capture from
        the \regex{} on the right hand side.  The list of acceptable
        variable names is:
